---
title: "Lab 5: Analysis of Variance"
author: "Elizabeth Miles-Flynn"
date: "2026-02-24"
output: pdf_document
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary packages
library(here)
library(tidyverse)
library(moments)
library(lterdatasampler)
```

# 1. Coastal Giant Salamander
```{r 1a test if salamander length and habitat have normal distribution}
# Filter for cascades (C), pools (P), and side channels (SC) for only Coastal giant salamander species. Plot and test whether or not there is normal distribution in dataset.
str(and_vertebrates)
view(and_vertebrates)

salamander_data <- and_vertebrates %>%
  filter(unittype %in% c("C", "P", "SC"),
         species == "Coastal giant salamander")

salamander_fig1 <- ggplot(salamander_data, 
                          aes(x = length_2_mm, fill = unittype)) +
  geom_histogram() +
    facet_wrap(~unittype) +
  scale_fill_manual(values = c("C" = "darkorange",
                               "P" = "purple",
                               "SC" = "cyan4")) +
  labs(x = "Snout-to-Tail Length (mm)",
       y = "Salamander Count",
       title = "Distribution of Coastal Giant Salamander Lengths Across Various 
       Habitat Types",
       caption = "Figure 1. Coastal Giant Salamanders in cascades (C), pools (P), 
       and side channel (SC) habitats. Dataset collected by H.J. Andrews Long 
       Term Ecological Research program within the Mack Creek watershed in Oregon") +
  theme(legend.position = "none")

salamander_fig1

Habitat_C <- salamander_data %>%
  filter(unittype == "C")
skew_C <- skewness(Habitat_C$length_2_mm, na.rm = TRUE)
kurt_C <- kurtosis(Habitat_C$length_2_mm, na.rm = TRUE)

Habitat_P <- salamander_data %>%
  filter(unittype == "P")
skew_P <- skewness(Habitat_P$length_2_mm, na.rm = TRUE)
kurt_P <- kurtosis(Habitat_P$length_2_mm, na.rm = TRUE)

Habitat_SC <- salamander_data %>%
  filter(unittype == "SC")
skew_SC <- skewness(Habitat_SC$length_2_mm, na.rm = TRUE)
kurt_SC <- kurtosis(Habitat_SC$length_2_mm, na.rm = TRUE)

salamander_data_log <- salamander_data %>%
  mutate(length_log10 = log10(length_2_mm))

Habitat_C <- salamander_data_log %>%
  filter(unittype == "C")
skew_C_log <- skewness(Habitat_C$length_log10, na.rm = TRUE)
kurt_C_log <- kurtosis(Habitat_C$length_log10, na.rm = TRUE)

Habitat_P <- salamander_data_log %>%
  filter(unittype == "P")
skew_P_log <- skewness(Habitat_P$length_log10, na.rm = TRUE)
kurt_P_log <- kurtosis(Habitat_P$length_log10, na.rm = TRUE)

Habitat_SC <- salamander_data_log %>%
  filter(unittype == "SC")
skew_SC_log <- skewness(Habitat_SC$length_log10, na.rm = TRUE)
kurt_SC_log <- kurtosis(Habitat_SC$length_log10, na.rm = TRUE)
```

```{r 1b Test if satisfy need for homogeneity of variances across groups}
salamander_var <- bartlett.test(salamander_data_log$length_log10,
                                salamander_data_log$unittype)
salamander_var
```

```{r 1c Test significant difference between salamander length and habitat types with a One-way ANOVA analysis}
salamander_ANOVA <- aov(length_log10 ~ unittype, data = salamander_data_log)

summary(salamander_ANOVA)

salamander_Tukey <- TukeyHSD(salamander_ANOVA)
salamander_Tukey
```

##### Question 1 Answers
1a. All habitats are not normally distribution with data strongly skewed to the right (0.5 > skewness value > 1.0, meaning moderately skewed) with heavy tails (kurtosis value > 3, meaning leptokurtic), which is displayed in Figure 1. After applying the log-transformation to the snout-to-length data, quick skewness and kurtosis values indicate that all three habitats closely meet the criteria for normal distribution (skewness around 0 AND kurtosis around 3 values).

1b. Reject the null hypothesis of Bartlett’s test that there is equal variances across groups because the p-value (0.003674) is much smaller than 0.05. In other words, variances are not homogeneous after applying the log-transformation for salamander length and habitat type. 

1c. Data from part b did not display approximately equal variances!


# 2. Coastal Giant Salamander
```{r 2a test if cutthroat trout length and reach have normal distribution}
# a. Filter for reach and cutthroat trout species. Plot and test whether or not there is normal distribution in dataset.

trout_data <- and_vertebrates %>%
  filter(species == "Cutthroat trout")
unique(trout_data$reach)


trout_fig2 <- ggplot(trout_data, 
                          aes(x = length_1_mm, fill = reach)) +
  geom_histogram() +
    facet_wrap(~reach) +
  scale_fill_manual(values = c("L" = "darkorange",
                               "M" = "purple",
                               "U" = "cyan4")) +
  labs(x = "Snout-to-Fork Length (mm)",
       y = "Trout Count",
       title = "Distribution of Cutthroat Trout Lengths Across Various Reach Types",
       caption = "Cutthroat Trout in Mack Creek Cut Forest lower reach (L), 
       middle reach (M), and upper reach (U) types. Dataset collected by H.J. 
       Andrews Long Term Ecological Research program within the Mack Creek watershed 
       in Oregon") +
  theme(legend.position = "none")

trout_fig2

Reach_L <- trout_data %>%
  filter(reach == "L")
skew_L <- skewness(Reach_L$length_1_mm, na.rm = TRUE)
kurt_L <- kurtosis(Reach_L$length_1_mm, na.rm = TRUE)

Reach_M <- trout_data %>%
  filter(reach == "M")
skew_M <- skewness(Reach_M$length_1_mm, na.rm = TRUE)
kurt_M <- kurtosis(Reach_M$length_1_mm, na.rm = TRUE)

Reach_U <- trout_data %>%
  filter(reach == "U")
skew_U <- skewness(Reach_U$length_1_mm, na.rm = TRUE)
kurt_U <- kurtosis(Reach_U$length_1_mm, na.rm = TRUE)
```

```{r 2b Test if satisfy need for homogeneity of variances across groups}
trout_var <- bartlett.test(trout_data$length_1_mm,
                                trout_data$reach)
trout_var
```

```{r 2c Test significant difference between trout length and reach types with a One-way ANOVA analysis}
trout_ANOVA <- aov(length_1_mm ~ reach, data = trout_data)

summary(trout_ANOVA)

trout_Tukey <- TukeyHSD(trout_ANOVA)
trout_Tukey

trout_means <- trout_data %>%
  group_by(reach) %>%
  summarise(mean_length = mean(length_1_mm, na.rm = TRUE))

trout_means

trout_fig3 <- ggplot(trout_means, 
                     aes(x = reach, y = mean_length, fill = reach)) +
  geom_col() +
  scale_fill_manual(values = c("L" = "darkorange",
                               "M" = "purple",
                               "U" = "cyan4")) +
  labs(x = "Reach Type",
       y = "Mean Snout-to-Fork Length (mm)",
       title = "Mean Cutthroat Trout Lengths Across Stream Reaches",
       caption = "Figure 3. Mean Cutthroat Trout lengths in Mack Creek Cut 
       Forest lower (L), middle (M), and upper (U) reaches. Dataset collected 
       by the H.J. Andrews Long Term Ecological Research program.") +
  theme(legend.position = "none")

trout_fig3

```

##### Question 2 Answers
2a. All reaches are meet the criteria for normal distribution with data slightly skewed to the right (0.4 > skewness value > 0, meaning moderately skewed yet closer to almost normal distribution) with very light tails (kurtosis value < 3, meaning slightly platykurtic), which is displayed in Figure 2.

2b. Reject the null hypothesis of Bartlett’s test that there is equal variances across groups because the p-value (0.004756) is much smaller than 0.05. In other words, variances in trout length are not homogeneous across different reaches. 

2c. There is evidence of significant differences between trout length and reach type because the p-value (2e‑16) is smaller than 0.05 and a large F-statistic (46.87; meaning variation between reaches are large and variation within reaches are small). Post‑hoc Tukey comparisons result with trout lengths similar between Uupper and Lower reaches (p = 0.584, approximately 0.59 mm differences in values). Middle reach were significantly larger than those from both U and L reaches (p < 0.000001 for both comparisons), meaning that Middle reach Cutthroat troat were generally larger in length.